Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-01 paper.qmd
revised figure 5

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, aes(x = tSNE1, y = tSNE2)) +
  geom_point(alpha = 0.5, color = "#636363", size = 0.5)

# First plot with xlim and ylim
nldr_two_curvy <- nldr_two_curvy_original+
  interior_annotation("a", sc_ltr_pos, cex = 1.5) +
  xlim(sc_xlims) +
  ylim(sc_ylims)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = tSNE1, y = tSNE2), 
             alpha = 0.5, size = 0.5, color = "#636363") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)
```

@@

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 1, 
               scaled_data = scaled_two_curvy_data, 
               scaled_data_model = scaled_two_curvy_data_model, 
               distance_df_small_edges = distance_df_small_edges_two_curvy, 
               axis_param = list(limits = 0.5, 
                                 axis_scaled = 2,
                                 axis_pos_x = -0.5, 
                                 axis_pos_y = -0.5, 
                                 threshold = 0.03))

@@

@@

two_curvy_proj_tsne_first_model1 <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.2, "#636363"), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    title = "c", 
    cex = 1.5, 
    position = sc_ltr_pos,
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(aspect.ratio = 1,
          legend.position = "none")

two_curvy_proj_true <- plot_proj(
  projected_df = projected_df, 
  model_df = true_model_df_proj, 
  axes = axes, 
  circle = circle, 
  point_param = c(0.5, 0.5, "#636363"), # size, alpha, color
  line_param = c(0.5, 0.5, "#2b8cbe"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  title = "a", 
  cex = 1.5, 
  axis_text_size = 4, 
  is_category = FALSE) +
  theme(aspect.ratio=1, 
        legend.position = "none")

two_curvy_proj_tsne_all_model1 <- plot_proj(
  projected_df = projected_df, 
  model_df = true_model_df_proj, 
  axes = axes, 
  circle = circle, 
  point_param = c(0.5, 0.5, "#636363"), # size, alpha, color
  line_param = c(0.5, 0.5, "#2b8cbe"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  title = "d", 
  cex = 1.5, 
  position = sc_ltr_pos,
  axis_text_size = 4, 
  is_category = FALSE) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000")

## Second projection
projection <- cbind(
  c(-0.38281,-0.04870,0.00619,-0.02451),
  c(-0.01303,0.00634,-0.37421,0.09648))

proj_obj2 <- get_projection(projection = projection, 
                            proj_scale = 1, 
                            scaled_data = scaled_two_curvy_data, 
                            scaled_data_model = scaled_two_curvy_data_model, 
                            distance_df_small_edges = distance_df_small_edges_two_curvy, 
                            axis_param = list(limits = 0.5, 
                                              axis_scaled = 6,
                                              axis_pos_x = -0.3, 
                                              axis_pos_y = -0.3, 
                                              threshold = 0))

@@

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% projection

@@

two_curvy_proj_true2 <- plot_proj(
    projected_df = projected_df, 
    model_df = true_model_df_proj, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.5, "#636363"), # size, alpha, color
    line_param = c(0.5, 0.5, "#2b8cbe"), # linewidth, alpha
    plot_limits = c(-0.42, 0.42), 
    title = "b", 
    cex = 1.5, 
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(aspect.ratio = 1, 
          legend.position = "none")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-tsne
#| fig-cap: "We start with a tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (b) and then lifted to $4\\text{-}D$. A \\gD{} projection of the model fit from a tour through the $4\\text{-}D$ space is presented in c. Additionally, a \\gD{} projection of the fitted and true models (d) from the same tour demonstrates that the model accurately captures the non-linear structure of the data, although it does not fully represent the width of that structure. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. The default is used. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

nldr_two_curvy2 + 
  trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 2)
```

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = "#0077A3",
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.05,
    color = "#0077A3") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = "#0077A3",
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    color = "#0077A3") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = "#0077A3",
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    color = "#0077A3") +
  geom_segment(
    data = model_df,
    aes(
      x = proj1_from,
      y = proj2_from,
      xend = proj1_to,
      yend = proj2_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2, label=rownames(axes)),
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)

>>>>

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = tSNE1, y = tSNE2)) +
  geom_point(alpha = 0.5, color = clr_choice, size = 0.5)

# First plot with xlim and ylim
nldr_two_curvy <- nldr_two_curvy_original +
  interior_annotation("a", sc_ltr_pos, cex = 1.5) +
  xlim(sc_xlims) +
  ylim(sc_ylims)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = tSNE1, y = tSNE2), 
             alpha = 0.5, size = 0.5, color = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)
```

@@

## First projection
model_prj1 <- cbind(
  c(0.37847,-0.06517,0.04231,0.01655),
  c(-0.06466,-0.27052,0.22131,-0.15235))

proj_obj1 <- get_projection(projection = model_prj1, 
               proj_scale = 1, 
               scaled_data = scaled_two_curvy_data, 
               scaled_data_model = scaled_two_curvy_data_model, 
               distance_df_small_edges = distance_df_small_edges_two_curvy, 
               axis_param = list(limits = 0.5, 
                                 axis_scaled = 2,
                                 axis_pos_x = -0.5, 
                                 axis_pos_y = -0.5, 
                                 threshold = 0.03))

@@

projection_scaled <- model_prj1 * 1

@@

two_curvy_proj_tsne_first_model1 <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.2, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    title = "c", 
    cex = 1.5, 
    position = sc_ltr_pos,
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(aspect.ratio = 1,
          legend.position = "none")

two_curvy_proj_true <- plot_proj(
  projected_df = projected_df, 
  model_df = true_model_df_proj, 
  axes = axes, 
  circle = circle, 
  point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  title = "a", 
  cex = 1.5, 
  axis_text_size = 4, 
  is_category = FALSE) +
  theme(aspect.ratio=1, 
        legend.position = "none")

## Second projection
model_prj2 <- cbind(
  c(-0.38281,-0.04870,0.00619,-0.02451),
  c(-0.01303,0.00634,-0.37421,0.09648))

proj_obj2 <- get_projection(projection = model_prj2, 
                            proj_scale = 1, 
                            scaled_data = scaled_two_curvy_data, 
                            scaled_data_model = scaled_two_curvy_data_model, 
                            distance_df_small_edges = distance_df_small_edges_two_curvy, 
                            axis_param = list(limits = 0.5, 
                                              axis_scaled = 6,
                                              axis_pos_x = -0.3, 
                                              axis_pos_y = -0.3, 
                                              threshold = 0))

@@

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% model_prj2

@@

two_curvy_proj_true2 <- plot_proj(
    projected_df = projected_df, 
    model_df = true_model_df_proj, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "black"), # linewidth, alpha
    plot_limits = c(-0.42, 0.42), 
    title = "b", 
    cex = 1.5, 
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(aspect.ratio = 1, 
          legend.position = "none")

two_curvy_proj_tsne_all_model1 <- plot_proj(
  projected_df = projected_df, 
  model_df = model_df, 
  axes = axes, 
  circle = circle, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  title = "d", 
  cex = 1.5, 
  position = sc_ltr_pos,
  axis_text_size = 4, 
  is_category = FALSE) #+
  #geom_segment(
  #  data = model_df, 
  #  aes(
  #    x = proj1_from, 
  #    y = proj2_from, 
  #    xend = proj1_to, 
  #    yend = proj2_to), 
  #  color = "#000000")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (b) and then lifted to $4\\text{-}D$ (c, d). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

nldr_two_curvy2 + 
  trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 4)
```

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = clr_choice,
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.05,
    color = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = clr_choice,
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    color = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = clr_choice,
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

five_gau_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    color = clr_choice) +
  geom_segment(
    data = model_df,
    aes(
      x = proj1_from,
      y = proj2_from,
      xend = proj1_to,
      yend = proj2_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2, label=rownames(axes)),
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)
